Programming Language Foundations in Agda

### 2019-07-12 papers/scp/PLFA.tex
additions to section 2, and spell check

Alas, there is no textbook for programming language theory in
Agda.  \emph{Verified Functional Programming in Agda} by \citep{Stump-2016}
covers related ground, but focusses more on programming with dependent
types than on the theory of programming languages.

@@

PLFA is aimed at students in the last year of an undergraduate
honours programme or the first year of a master or doctorate degree.
It aims to teach the fundamentals of operational semantics of
programming languages, with simply-typed lambda calculus as the
central example.  The textbook is written as a literate script in Agda.
As with SF, the hope is that using
a proof assistant will make the development more concrete
and accessible to students, and give them rapid feedback to find
and correct misaprehensions.

@@

\paragraph{Big-step: evaluation for call-by-name}
Introduces the notion of big-step evaluation to develop a
deterministic call-by-name reduction strategy. The main result of this
chapter is a proof that big-step evaluation implies the existence of a
reduction sequence that terminates with a lambda abstraction.

@@

\paragraph{Denotational: semantics of the untyped lambda calculus}
The early denotational semantics of the lambda calculus based on graph
models~\citep{Scott-1976,Engeler-1981,Plotkin-1993} and filter
models~\citep{Barendregt-1983} were particularly simple and elegant.
This chapter presents such a semantics in a big-step style that is
approachable to readers familiar with operational semantics.

\paragraph{Compositional: the denotational semantics is compositional}
TODO

\paragraph{Soundness: of reduction with respect to denotational semantics}
TODO

\paragraph{Adequacy: of denotational semantics with respect to reduction}
TODO

\paragraph{Contextual Equivalence: is implied by denotational equality}
TODO

@@

The second part of PLFA first discusses two different approaches to
modelling simply-typed lambda calculus.  It first presents raw
terms with named variables and a separate typing relation and
then shifts to inherently-typed terms with de Bruijn indices.
Before writing the text, I had thought the two approaches
complementary, with no clear winner.  Now I am convinced that the
inherently-typed approach is superior.

Figure~\ref{fig:raw} presents the raw approach.
It first defines $\texttt{Id}$, $\texttt{Term}$,
$\texttt{Type}$, and $\texttt{Context}$, the abstract syntax
of identifiers, raw terms, types, and contexts.
It then defines two judgements,
$\GG\:{\ni}\:\ex\:{\typecolon}\:\AY$ and
$\GG\:{\vdash}\:\EM\:{\typecolon}\:\AY$,
which hold when under context $\GG$ the variable $\ex$
and the term $\EM$ have type $\AY$, respectively.

Figure~\ref{fig:inherent} presents the inherent approach.
It first defines $\texttt{Type}$ and $\texttt{Context}$, the abstract syntax
of types and contexts, of which the first is as before and the second is
as before with identifiers dropped.  In place of the two judgements,
the types of variables and terms are indexed by a context and a type,
so that $\GG\:{\ni}\:\AY$ and $\GG\:{\vdash}\:\AY$ denote
variables and terms, respectively, that under context $\GG$ hae type $\AY$.
The indexed types closely resemble the previous judgements:
we now represent a variable or a term by the proof that it is well-typed.
In particular, the proof that a variable is well-typed in the raw approach
corresponds to a de Bruijn index in the inherent approach.

The raw approach requires more lines of code than
the inherent approach.  The separate definition of raw
terms is not needed in the inherent
approach; and one judgement in the raw approach needs to check that
$\ex \not\equiv \why$, while the corresponding judgement in the
inherent approach does not.  The difference becomes more pronounced
when including the code for substitution, reductions, and proofs of
progress and preservation.  In particular, where the raw approach
requires one first define substitution and reduction and then prove
they preserve types, the inherent approach establishes substitution at
the same time it defines substitution and reduction.

@@

There are actually four possible designs, as the choice of named
variables vs de Bruijn indices, and the choice of raw vs
inherently-typed terms may be made independently.  There are synergies
beween the two.  Manipulation of de Bruijn indices can be notoriously
error-prone without inherent-typing to give assurance of correctness.
In inherent typing with named variables, simultaneous substitution by
open terms remains difficult.

@@

\paragraph{Acknowledgements}

@@

>>>>

Alas, there is no textbook for programming language theory in
Agda.  \emph{Verified Functional Programming in Agda} by \citep{Stump-2016}
covers related ground, but focuses more on programming with dependent
types than on the theory of programming languages.

@@

PLFA is aimed at students in the last year of an undergraduate
honours programme or the first year of a master or doctorate degree.
It aims to teach the fundamentals of operational semantics of
programming languages, with simply-typed lambda calculus as the
central example.  The textbook is written as a literate script in Agda.
As with SF, the hope is that using
a proof assistant will make the development more concrete
and accessible to students, and give them rapid feedback to find
and correct misapprehensions.

@@

\paragraph{Big-step: evaluation for call-by-name}
Introduces the notion of big-step evaluation, written $\gamma \vdash M
\Downarrow V$, to develop a deterministic call-by-name reduction
strategy. The main result of this chapter is a proof that big-step
evaluation implies the existence of a reduction sequence that
terminates with a lambda abstraction.

@@

\paragraph{Denotational: semantics of the untyped lambda calculus}
The early denotational semantics of the lambda calculus based on graph
models~\citep{Scott-1976,Engeler-1981,Plotkin-1993} and filter
models~\citep{Barendregt-1983} were particularly simple and elegant: a
function is modeled as a lookup table.  This chapter presents such a
semantics using a big-step notation that is approachable to readers
familiar with operational semantics, writing $\gamma \vdash M
\downarrow d$ for the evaluation of a term $M$ to denotation $d$ in
environment $\gamma$.

\paragraph{Compositional: the denotational semantics is compositional}
The hallmark of denotational semantics is that they are compositional:
the meaning of each language form is a function of the meanings of its
parts.
%
We define two functions, named \texttt{curry} and \texttt{apply} that
serve this purpose for lambda abstraction and application.
%
The results in this chapter include congruences for \texttt{curry} and
\texttt{apply} and the compositionality property for the denotational
semantics.
%
The chapter concludes with a functional definition of the denotational
semantics and a proof that it is equivalent to the big-step version.

\paragraph{Soundness: of reduction with respect to denotational semantics}
Reduction implies denotational equality. We prove each direction of
the equality, first showing the reduction preserves denotations
(subject reduction), and then showing that reduction reflects
denotations (subject expansion). The first proof is similar to the
type preservation proofs in Part II. The second goes in reverse,
showing that if $M \longrightarrow N$ and $\gamma \vdash N
\downarrow d$, then $\gamma \vdash M \downarrow d$ .

\paragraph{Adequacy: of denotational semantics with respect to reduction}
If a term is denotationally equal to a lambda expression, then it
reduces to a lambda expression. The main lemma shows that if a term's
denotation is functional, i.e., $\gamma \vdash M \downarrow (d \mapsto
d')$, then $M$ terminates according to the call-by-name big-step
semantics, i.e., $\gamma' \vdash M \Downarrow V$. A logical relation
$\mathbb{V}$ is used to relate denotations and values (i.e. closures).
The implication from the big-step semantics to reduction is proved in
the Big-step chapter of Part II.

\paragraph{Contextual Equivalence: is implied by denotational equality}
The main criteria for behavior-preserving code transformation (such as
compiler optimization or programmer refactoring) is contextual
equivalence. Two terms are contextually equivalent when they can both
be placed into an arbitrary context (a program with a hole) and the
resulting programs behave the same (e.g., they both terminate or they
both diverge). This chapter ties together the previous results
(Compositionality, Soundness, and Adequacy) to show that denotational
equality implies contextual equivalence. Thus, it is safe to use
denotational equality to justify code transformations.

@@

The second part of PLFA first discusses two different approaches to
modeling simply-typed lambda calculus.  It first presents raw
terms with named variables and a separate typing relation and
then shifts to inherently-typed terms with de Bruijn indices.
Before writing the text, I had thought the two approaches
complementary, with no clear winner.  Now I am convinced that the
inherently-typed approach is superior.

Figure~\ref{fig:raw} presents the raw approach.
It first defines $\texttt{Id}$, $\texttt{Term}$,
$\texttt{Type}$, and $\texttt{Context}$, the abstract syntax
of identifiers, raw terms, types, and contexts.
It then defines two judgments,
$\GG\:{\ni}\:\ex\:{\typecolon}\:\AY$ and
$\GG\:{\vdash}\:\EM\:{\typecolon}\:\AY$,
which hold when under context $\GG$ the variable $\ex$
and the term $\EM$ have type $\AY$, respectively.

Figure~\ref{fig:inherent} presents the inherent approach.
It first defines $\texttt{Type}$ and $\texttt{Context}$, the abstract syntax
of types and contexts, of which the first is as before and the second is
as before with identifiers dropped.  In place of the two judgments,
the types of variables and terms are indexed by a context and a type,
so that $\GG\:{\ni}\:\AY$ and $\GG\:{\vdash}\:\AY$ denote
variables and terms, respectively, that under context $\GG$ have type $\AY$.
The indexed types closely resemble the previous judgments:
we now represent a variable or a term by the proof that it is well-typed.
In particular, the proof that a variable is well-typed in the raw approach
corresponds to a de Bruijn index in the inherent approach.

The raw approach requires more lines of code than
the inherent approach.  The separate definition of raw
terms is not needed in the inherent
approach; and one judgment in the raw approach needs to check that
$\ex \not\equiv \why$, while the corresponding judgment in the
inherent approach does not.  The difference becomes more pronounced
when including the code for substitution, reductions, and proofs of
progress and preservation.  In particular, where the raw approach
requires one first define substitution and reduction and then prove
they preserve types, the inherent approach establishes substitution at
the same time it defines substitution and reduction.

@@

There are actually four possible designs, as the choice of named
variables vs de Bruijn indices, and the choice of raw vs
inherently-typed terms may be made independently.  There are synergies
between the two.  Manipulation of de Bruijn indices can be notoriously
error-prone without inherent-typing to give assurance of correctness.
In inherent typing with named variables, simultaneous substitution by
open terms remains difficult.

@@

\paragraph{Acknowledgments}

@@

% LocalWords:  Agda adr Kokke Siek Wadler Woodlawn Bloomington Coq et
% LocalWords:  PLFA evaluator normalise programme programmes al Huet
% LocalWords:  ICFP subst Bove Mixfix de Janeiro summarising favour
% LocalWords:  organised formalising Colour Gonthier CompCert Kastner
% LocalWords:  SEL honours formalisms monus associativity reflexivity
% LocalWords:  commutativity equational of's Isomorphism isomorphism
% LocalWords:  extensionality disjunction intuitionistic existentials
% LocalWords:  Booleans booleans monoid specialisation PCF Plotkin EP
% LocalWords:  fixpoint DeBruijn Bruijn indices Bisimulation Untyped
% LocalWords:  bisimulation normalisation untyped Abadi Barendregt
% LocalWords:  Rosser Tait Schafer Tebbi Smolka Pfenning Denotational
% LocalWords:  denotational Engeler lookup Compositional congruences
% LocalWords:  compositional compositionality denotationally Hoare
% LocalWords:  refactoring subtyping Appel QuickChick formalise SF's
% LocalWords:  Stlc StlcProp Coq's reinvoke Agda's postconditions Ulf
% LocalWords:  flavour layed emphasises tvar tapp CoqIDE sublanguage
% LocalWords:  submenus Danas monospaced FreeMono Felleisen Redex
% LocalWords:  Rosu Serbanuta realise Capretta cryptocurrencies Oleg
% LocalWords:  evaluators Dagand Scherer Kiselyov eval Goguen McKinna
% LocalWords:  Altenkirch Reus Allais Darais Conor Norell Andreas
% LocalWords:  EPSRC
